{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-08 thesis/related.tex
write intro/motivation for htm
weak atomicity description, proof, & implementation, wee

Testing approaches for transactional programs are sparsely represented in the literature so far.
Although several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm}
are building up to formal proofs of the correctness of underlying TM {\em implementations},
Landslide is the first I know of to verify client programs thereof.
% TODO: citation as noun - global scan
\revision{\cite{mc-tm-with-spin} used SPIN \cite{spin} to model check McRT STM \cite{mcrt-stm},
an STM implementation built by Intel,
verifying}
up to 2 threads running 1 transaction each with up to 3 memory accesses.
This kind of verification,
analogous to \cref{sec:education-pebbles-tests}'s {\tt mutex\_test},
is an important stepping stone for trusting the results Landslide will provide.
STAMP \cite{stamp} is a benchmark suite transactional programs,
implemented using the OpenTM interface \cite{opentm},
used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
As discussed in \cref{sec:tm-eval-exp-setup},
it focuses more on performance than on interesting concurrency properties.
% stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
% stamp need not be considered harmful put here
Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
and that scalable HTM programs must
minimize incidental conflicts and handle aborts more flexibly than with blind retry loops.
The programming complexity needed to achieve these goals calls,
of course,
for correspondingly advanced verification approaches such as Landslide.
Finally, TxRace \cite{txrace}
tests non-transactional programs for data races
by inserting HTM calls via compiler instrumentation,
% they don't actually seem to check the failure reason lmao
% i'll phrase this in a slightly weasel way to point out that's what they *should* be doing
% and that in turn motivates the stuff i did
% even tho they use a diff way to filter them post hoc (well, they'd still need to, bc of false sharing)
relying on conflict aborts to point out access pairs that would be unsafe in the original program.
This citation arguably belongs in \cref{sec:related-data-race} as well;
I include it here to highlight the importance of Landslide's
ability to distinguish different abort reasons (\cref{sec:txn-abort-codes}).

An article from relatively early in the timeline of TSX \cite{htm-subtleties}
warns of several false equivalence pitfalls when converting conventionally-locking code to use transactions,
although these pitfalls depend on multiple existing locks used locally and disjointly,
so this does not invalidate the equivalence proved in \cref{sec:tm-design-locks}.
Rather, Landslide could be used to ensure that freshly-converted transactional code avoids the warned-of pitfalls.
{\em Learning from Mistakes} \cite{learning-from-mistakes},
a survey of the characteristics of many types of concurrency bugs,
found that TM could potentially fix some, but not all, of the studied bugs,
while in other cases it must be combined with other concurrency primitives to be fully correct.
A subsequent paper \cite{applying-tm-bugs}
found a majority of bugs in their study to be easily fixable with hand-written transactions,
while others remained out of scope due to blocking {\tt cond\_wait()} operations and the like;
more recently, the tool BugTM \cite{bugtm} aims to deploy such repairs in production code fully automatically.
However, these studies all optimize for empirical correctness at best,
as well as maintaining good performance,
which motivates the use of tools like Landslide to ensure these rewrites
are actually correct,
% oops
rather than merely shrinking the necessary preemption window required to expose them.

@@

\subsubsection{Symbolic execution}

>>>>

Testing approaches for transactional programs are sparsely represented in the literature so far.
Although several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm}
are building up to formal proofs of the correctness of underlying TM {\em implementations},
Landslide is the first I know of to verify client programs thereof.
% TODO: citation as noun - global scan
\revision{McRT STM \cite{mcrt-stm},
an STM implementation built by Intel,
was checked using SPIN \cite{spin}, verifying
up to 2 threads running 1 transaction each with up to 3 memory accesses \cite{mc-tm-with-spin}.}
This kind of verification,
analogous to \cref{sec:education-pebbles-tests}'s {\tt mutex\_test},
is an important stepping stone for trusting the results Landslide will provide.
STAMP \cite{stamp} is a benchmark suite transactional programs,
implemented using the OpenTM interface \cite{opentm},
used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
As discussed in \cref{sec:tm-eval-exp-setup},
it focuses more on performance than on interesting concurrency properties.
% stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
% stamp need not be considered harmful put here
Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
and that scalable HTM programs must
minimize incidental conflicts and handle aborts more flexibly than with blind retry loops.
The programming complexity needed to achieve these goals calls,
of course,
for correspondingly advanced verification approaches such as Landslide.
Finally, TxRace \cite{txrace}
tests non-transactional programs for data races
by inserting HTM calls via compiler instrumentation,
% they don't actually seem to check the failure reason lmao
% i'll phrase this in a slightly weasel way to point out that's what they *should* be doing
% and that in turn motivates the stuff i did
% even tho they use a diff way to filter them post hoc (well, they'd still need to, bc of false sharing)
relying on conflict aborts to point out access pairs that would be unsafe in the original program.
This citation arguably belongs in \cref{sec:related-data-race} as well;
I include it here to highlight the importance of Landslide's
ability to distinguish different abort reasons (\cref{sec:txn-abort-codes}).

%An article from relatively early in the timeline of TSX
\revisionminor{The paper which originally defined weak and strong atomicity~\cite{htm-subtleties}
also}
warns of several false equivalence pitfalls when converting conventionally-locking code to use transactions,
although these pitfalls depend on multiple existing locks used locally and disjointly,
so this does not invalidate the equivalence proved in \cref{sec:tm-design-locks}.
Rather, Landslide could be used to ensure that freshly-converted transactional code avoids the warned-of pitfalls.
{\em Learning from Mistakes} \cite{learning-from-mistakes},
a survey of the characteristics of many types of concurrency bugs,
found that TM could potentially fix some, but not all, of the studied bugs,
while in other cases it must be combined with other concurrency primitives to be fully correct.
A subsequent paper \cite{applying-tm-bugs}
found a majority of bugs in their study to be easily fixable with hand-written transactions,
while others remained out of scope due to blocking {\tt cond\_wait()} operations and the like;
more recently, the tool BugTM \cite{bugtm} aims to deploy such repairs in production code fully automatically.
However, these studies all optimize for empirical correctness at best,
as well as maintaining good performance,
which motivates the use of tools like Landslide to ensure these rewrites
are actually correct,
% oops
rather than merely shrinking the necessary preemption window required to expose them.

@@

\subsubsection{Symbolic execution}
\label{related:symbolic}

### 2018-12-08 thesis/tm.tex
weak atomicity description, proof, & implementation, wee

% do they track changes in pthread-style TLS? or?
{\bf Implementations.}
Software TM implementations (STM) typically function as a library,
tracking staged memory accesses in local memory,
and aborting whenever a conflict is detected between two transactions' tracked accesses \cite{stm-pldi06}.
Hardware TM implementations (HTM) use processor-level hardware support,
which stages changes in per-CPU cache lines,
and may abort for STM's reason above \cite{intel-tsx-overview},
%\cite{htm-experience},
or additionally whenever a conflict is detected between one transaction's traced access and {\em any} other memory access,
or whenever a conflict occurs on the same cache line, not necessarily the same address,
or in case of any system interrupt or cache overflow.
Intel's commercial implementation of HTM has a rocky history of hardware bugs \cite{intel-disable-tsx,intel-tsx-update},
attesting to the feature's complexity and the need for formal verification on both sides of its API.

{\bf Terminology.}
The world of hardware transactional memory is home to several more confusing acronyms in addition to ``HTM''.
Transactional Synchronization Extensions (TSX)
refers to Intel's implementation of HTM on Haswell and more recent microarchitectures
\cite{htm-haswell}.
Restricted Transactional Memory (RTM)
refers to the {\tt xbegin}, {\tt xend}, and {\tt xabort} subset of TSX instructions,
which of course correspond to {\sf begin}, {\sf end}, and {\sf abort} listed above,
as well as {\tt xtest}, an instruction which returns whether or not the CPU is executing transactionally.
GCC and Clang expose these as C/C++ intrinsics named {\tt \_xbegin()}, {\tt \_xend()}, {\tt \_xabort()}, and {\tt \_xtest()}
\cite{htm-gcc}.
Hardware Lock Elision (HLE)
refers to the {\tt xacquire} and {\tt xrelease} subset of TSX instructions,
which extend the traditional interface to offer a slightly higher-level way to access the CPU feature,
optimized for simplicity for locking-like synchronization patterns
% other possible things to cite here:
% https://lwn.net/Articles/534758/
% https://software.intel.com/en-us/node/683688
\cite{speculative-lock-elision,lock-elision}.
In this thesis I focus on RTM, the more general (i.e., expressive (i.e., bug-prone)) interface,
and among all these acronyms restrict myself to ``HTM''
(when referring to transactional memory as a concurrency primitive in the abstract)
and ``TSX''
(when referring to Intel's implementation and/or GCC's intrinsics interface).
The non-pedantic reader may treat these as interchangeable.

% or "Contribution"
\revision{{\bf Motivation.}
For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
addresses input nondeterminism; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

Then, without loss of generality (for any number of other threads \tjj/\tkk, and for any number of thread switches away from \tii during the transaction):
\vspace{1em}

@@

Let $\tiiat{\mu},\tjjat{\nu}$ be a pair of memory accesses to the same address, at least one a write,
in some transactional execution $I$ normalized under \Cref{lem:equiv}.
Then let $\mathsf{lockify}_m(\tkkat{L})$ denote a function over instructions in $I$,
which replaces $\tkkat{L}$ with $\tkkat{\mathsf{lock}(m)}$ if $L$ is a successful HTM begin,
with a no-op if $L$ is a transaction abort,
or with $\tkkat{\mathsf{unlock}(m)}$ if $L$ is an HTM end,
or no replacement otherwise.
Finally, let $I' = \exists m. \mathsf{lockify}_m(I)$,
the execution with the boundaries of all successful transactions replaced by an abstract global lock.
\Cref{lem:equiv} guarantees mutual exclusion of $m$.
\vspace{1em}

\begin{theorem}[Transactions are a Global Lock]
	$\tiiat{\mu},\tjjat{\nu}$ is a data race in $I$ iff it is a data race in $I'$.
\end{theorem}

\begin{proof}
I prove one case for each variant definiton for data races supported in Landslide \cite{quicksand}.
For each, I state below what it means to race in an execution with synchronizing HTM instructions.

@@

@@

Support for TSX programs in Landslide is implemented in four parts, broadly speaking:
the user interface,
plus one internal part corresponding to each proof above,
plus a bonus optimization for pruning equivalent interleavings under the new concurrency model.

\subsection{User Interface}

@@

Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
%Each option requires each of the previously-listed options to also be present.
\begin{itemize}
	\item {\tt -X} (for ``tsX'' or ``Xbegin'') enables the basic features:
		preemption points on each {\tt \_xbegin()} and {\tt \_xend()} call,
		failure injection on each of the former
		(always returning {\tt \_XABORT\_RETRY} as the failure code)
		(\revisionminor{\cref{sec:tm-design-failure}}, \cref{sec:txn-failure}),
		and treatment of transactional regions during data-race analysis (\cref{sec:txn-datarace}).
	\item {\tt -A} (for ``xAbort codes'')
		enables multiple xabort failure codes (\cref{sec:txn-failure}).
		Requires {\tt -X}.
	\item {\tt -S} (for ``Stm'' or ``Suppress retries'')
		disables the {\tt \_XABORT\_RETRY} failure reason,
		causing Landslide to emulate the semantics of STM rather than HTM.
		%(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
		Requires {\tt -X -A}.
	\item {\tt -R} (for ``Retry sets'')
		enables state space reduction based on independences between individual {\tt xbegin} results
		(\cref{sec:tm-retrysets}).
		Requires {\tt -X} and {\em not} {\tt -A}.
\end{itemize}

@@

>>>>

\subsubsection{Implementations}

% do they track changes in pthread-style TLS? or?
Software TM implementations (STM) typically function as a library,
tracking staged memory accesses in local memory,
and aborting whenever a conflict is detected between two transactions' tracked accesses \cite{stm-pldi06,mcrt-stm,swisstm}.
Hardware TM implementations (HTM) use processor-level hardware support,
which stages changes in per-CPU cache lines,
and may abort for STM's reason above \cite{intel-tsx-overview},
%\cite{htm-experience},
or additionally whenever a conflict is detected between one transaction's traced access and {\em any} other memory access,
or whenever a conflict occurs on the same cache line, not necessarily the same address,
or in case of any system interrupt or cache overflow.
Intel's commercial implementation of HTM has a rocky history of hardware bugs \cite{intel-disable-tsx,intel-tsx-update},
attesting to the feature's complexity and the need for formal verification on both sides of its API.

\subsubsection{Terminology}

The world of transactional memory is home to several more confusing acronyms similar to ``HTM''.
Transactional Synchronization Extensions (TSX)
refers to Intel's implementation of HTM on Haswell and more recent microarchitectures
\cite{htm-haswell}.
Restricted Transactional Memory (RTM)
refers to the {\tt xbegin}, {\tt xend}, and {\tt xabort} subset of TSX instructions,
which of course correspond to {\sf begin}, {\sf end}, and {\sf abort} listed above,
as well as {\tt xtest}, an instruction which returns whether or not the CPU is executing transactionally.
GCC and Clang expose these as C/C++ intrinsics named {\tt \_xbegin()}, {\tt \_xend()}, {\tt \_xabort()}, and {\tt \_xtest()}
\cite{htm-gcc}.
Hardware Lock Elision (HLE)
refers to the {\tt xacquire} and {\tt xrelease} subset of TSX instructions,
which extend the traditional interface to offer a slightly higher-level way to access the CPU feature,
optimized for simplicity for locking-like synchronization patterns
% other possible things to cite here:
% https://lwn.net/Articles/534758/
% https://software.intel.com/en-us/node/683688
\cite{speculative-lock-elision,lock-elision}.
In this thesis I focus on RTM, the more general (i.e., expressive (i.e., bug-prone)) interface,
and among all these acronyms restrict myself to ``HTM''
(when referring to transactional memory as a concurrency primitive in the abstract)
and ``TSX''
(when referring to Intel's implementation and/or GCC's intrinsics interface).
The non-pedantic reader may treat these as interchangeable.

\revision{Transactional atomicity comes in two flavours,
{\em strong} and {\em weak atomicity}~\cite{htm-subtleties}.
%
HTM implementations use the processor's cache coherence protocol to detect conflicts,
and hence any conflicting memory access from another CPU will cause an abort,
regardless of whether that access was itself transactional.
Hence, HTM transactions are {\em strongly atomic}, i.e., appear indivisible to all other code.
%
STM implementations rely on the programmer to use its interface to access any shared memory
she intends to protect with its transactions,
which cannot detect conflicts with non-transactional accesses that bypass it.
Hence, STM transactions are {\em weakly atomic}, i.e.,
its transactions will abort only from conflicts with other transactions,
while other code may interleave with them freely.
%
The manner in which non-transactional code may interleave with STM transactions
further depends on whether the STM uses
{\em eager versioning},
wherein transactional updates are immediately visible to non-transactional code,
or
{\em lazy versioning},
wherein such updates are visible only after commit
\cite[\S{}2]{stm-weak-atomicity}.
Unless explicitly noted as weak atomicity,
this chapter assumes strong atomicity,
and ``STM semantics'' refers to a strongly-atomic HTM program emulating STM with retry loops
(see \cref{sec:txn-abort-codes}).
Treatment of weak atomicity in
\cref{sec:tm-weak-atomicity} and \cref{sec:tm-impl-weak-atomicity}
assumes eager versioning.}

% or "Contribution"
\subsubsection{Motivation}

\revision{For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
addresses input nondeterminism; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend the proofs to account for weak atomicity (\cref{sec:tm-weak-atomicity}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

Then, without loss of generality (for any number of other threads \tjj/\tkk,
\revisionminor{WLOG presented here as two threads,}
and for any number of thread switches away from \tii during the transaction,
\revisionminor{WLOG presented here as one switch}):
\vspace{1em}

@@

Let $\tiiat{\mu},\tjjat{\nu}$ be a pair of memory accesses to the same address, at least one a write,
in some transactional execution $I$ normalized under \Cref{lem:equiv}.
Then let $\mathsf{lockify}_m(\tkkat{L})$ denote a function over instructions in $I$,
which replaces $\tkkat{L}$ with $\tkkat{\mathsf{lock}(m)}$ if $L$ is a successful HTM begin,
with a no-op if $L$ is a transaction abort,
or with $\tkkat{\mathsf{unlock}(m)}$ if $L$ is an HTM end,
or no replacement otherwise.
Finally, let $I' = \exists m. \mathsf{lockify}_m(I)$,
the execution with the boundaries of all successful transactions replaced by an abstract global lock $m$.
\Cref{lem:equiv} guarantees mutual exclusion of $m$.
\vspace{1em}

\begin{theorem}[Transactions are a Global Lock]
	\label{thm:global-lock}
	$\tiiat{\mu},\tjjat{\nu}$ is a data race in $I$ iff it is a data race in $I'$.
\end{theorem}

\begin{proof}
I prove one case \revisionminor{for each of the two variant definitons of data races that Landslide supports
(\cref{sec:landslide-datarace})}.
%\cite{quicksand}. % wtf no
For each, I state below what it means to race in an execution with synchronizing HTM instructions.

@@

\subsection{Weak atomicity}
\label{sec:tm-weak-atomicity}

\newcommand\weakhilight[1]{{\ul{#1}}\xspace}

\revision{Prior work has highlighted the difference between
{\em strong} and {\em weak atomicity}
in TM systems
\cite{htm-subtleties,stm-weak-atomicity}.
Thus far we have assumed transactions will abort from any intervening memory conflict,
whether or not that access is itself transactional.
This corresponds to strong atomicity, the execution semantics provided by TSX.
% \cite{bugtm}. % sec 3.1 of that paper, not sure of an earlier citation that also states it in these exact terms
However, under weakly-atomic STM systems,
transactions do not abort when
a non-transactional conflicting access interleaves within them,
and condition \#1 of the execution semantics from \cref{sec:tm-design-failure} does not hold.
To model weak atomicity, we must weaken \Cref{lem:equiv} as follows (updated parts underlined for emphasis).
\vspace{1em}

\begin{lemma}[Equivalence of Aborts \weakhilight{under Weak Atomicity}]
	\label{lem:equiv-weak}
	For any interleaving prefix
	\[
	\begin{tabular}{c}
		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
	\end{tabular}
	\]
	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
	\begin{enumerate}
		\item \weakhilight{All events in
			$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},
			 \tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			 % ehh, bad parallel structure with below regarding the verb 'exists'
			are non-transactional, and the original interleaving itself exists, or}
		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
			(conflicting case), or
		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			(independent case)
	\end{enumerate}
	exists and is observationally equivalent.
\end{lemma}

\begin{proof}
	As before, \weakhilight{except first case on
	whether any intervening operations by $\tjj$/$\tkk$ are transactional.%
	\footnote{If $\tjj$/$\tkk$ already have a transaction active,
	their {\sf begin} must have occurred before $\tiiat{\alpha}$,
	and the theorem can apply starting from the earlier {\tt begin} with the thread IDs swapped.}
	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$ without aborting,
	and the new case \#1 applies.}
\end{proof}

\Cref{thm:atom} must be weakened correspondingly.
\vspace{1em}

\begin{theorem}[\weakhilight{Weak} Atomicity of Transactions]
	\label{thm:atom-weak}
	For any state space $S$ of a transactionally-concurrent program,
	an equivalent state space exists in which all transactions are either executed atomically
	\weakhilight{with respect to other transactions only},
	or aborted immediately.
\end{theorem}

\begin{proof}
	As before.
\end{proof}

Conveniently, in \cref{sec:tm-design-locks}'s formulation of
$I' = \exists m. \mathsf{lockify}_m(I)$,
\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
so mutual exclusion of $m$ is preserved,
%(a global lock does not protect against interleaving by arbitrary other code that never takes the lock either)
and
%the proof of
\Cref{thm:global-lock} applies unmodified.
\cref{sec:tm-impl-weak-atomicity}
discusses the implementation consequences of this weakening.}

@@

Support for TSX programs in Landslide is implemented in \revisionminor{five} parts, broadly speaking:
the user interface,
plus one internal part corresponding to each proof above,
plus a bonus optimization for pruning equivalent interleavings under the new concurrency model.

\subsection{User Interface}
\label{sec:tm-interface}

@@

Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
%Each option requires each of the previously-listed options to also be present.
\begin{itemize}
	\item {\tt -X} (for ``tsX'' or ``Xbegin'') enables the basic features:
		preemption points on each {\tt \_xbegin()} and {\tt \_xend()} call,
		failure injection on each of the former
		(always returning {\tt \_XABORT\_RETRY} as the failure code)
		(\revisionminor{\cref{sec:tm-design-failure}}, \cref{sec:txn-failure}),
		and treatment of transactional regions during data-race analysis (\cref{sec:txn-datarace}).
	\item {\tt -A} (for ``xAbort codes'')
		enables multiple xabort failure codes (\cref{sec:txn-failure}).
		Requires {\tt -X}.
	\item {\tt -S} (for ``Stm'' or ``Suppress retries'')
		disables the {\tt \_XABORT\_RETRY} failure reason,
		causing Landslide to emulate the semantics of STM rather than HTM.
		%(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
		Requires {\tt -X -A}.
	\item {\tt -R} (for ``Retry sets'')
		enables state space reduction based on independences between individual {\tt xbegin} results
		(\cref{sec:tm-retrysets}).
		Requires {\tt -X} and {\em not} {\tt -A}.
	\item \revision{{\tt -W} (for ``Weak atomicity'')
		enables weak atomicity, %for transactions,
		allowing non-transactional code to interleave within transactions,
		corresponding to \cref{sec:tm-weak-atomicity}.
		Requires {\tt -S}.}
\end{itemize}

@@

\subsection{Weak atomicity}
\label{sec:tm-impl-weak-atomicity}

\revision{If the user supplies the {\tt -W} option (\cref{sec:tm-interface}),
Landslide will emulate STM's weak atomicity semantics (\cref{sec:tm-weak-atomicity}).
%to allow non-transactional code
%to interleave between a {\tt \_xbegin()}/{\tt \_xend()} pair.
Concretely, it achieves this by
reporting both halves of data races that occur between transactional and non-transactional code (\cref{sec:landslide-datarace})
(previously, only the latter half would be reported),
allowing preemption points to be identified during transactions (\cref{sec:landslide-pps}),
and allowing system calls during transactions to proceed rather than forcing a {\tt CAPACITY} abort
(as programmers may now need to nest other synchronization within transactions to protect against non-transactional code).
%and traversing the interleaving arbitrarily far back from a {\tt \_xend()} to find the appropriate preemption
%(previously, the xbegin would be at the immediately previous PP)
Landslide requires the other STM options ({\tt -A -S}) also to be enabled to emulate weak atomicity,
because I know of no existing HTM system which implements it.
As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
and moreover,
every single HTM benchmark in the upcoming evaluation,
implemented to rely on strong atomicity,
exhibits bugs under weak atomicity.
Commit 68401fc implements this mode.}
